---
title: "Cohorts in R"
output: html_document
---

Cohorts are a fundamental building block of analytics using OHDSI tools. What are some things we can do with cohorts in R?

Recall that the OHDSI definition of a cohort is "A set of persons satisfying one or more inclusion criteria for a duration of time".

A "cohort definition"  is the set of rules that defines a cohort. This set of rules can be represented in multiple ways:

- Human readable text
- A json file
- SQL code

Each of these ways of representing a cohort definition are interchagable. However the json representation can be translated into the other two automatically by OHDSI tools so we will consider json as the primary representation of a cohort definition.

Atlas is a fantastic tool for creating cohort definitions. We can use R to manipulate these definitions after they are created. The basic things we might want to do with a cohort definiton are

- Save them for later use or to share with others
- Load them into Atlas programmatically
- Instantiate/Generate them on a specific database


We can think of every cohort as progressing through three phases.

```{r, echo=F}
DiagrammeR::grViz("digraph {

graph [layout = dot, rankdir = LR]

node [shape = rectangle, style = filled, fillcolor = Linen]

json [label =  'JSON: \n cohort \n definition']
sql [label = 'SQL: \n needed to \n generate cohort']
cohortTable [label= 'Cohort Table: \n the cohort subject IDs, \n entry, and exit dates']

json -> sql -> cohortTable
}")
```

The OHDSI tools help us move from one phase to the next.


# Download a cohort definition created in Atlas

The ROhdsiWebApi package gives us a way to programmatically interact with WebAPI (the Atlas backend). 


```{r}
library(ROhdsiWebApi)
baseUrl <- "https://eco.ohdsitutorials.amazingawsdemos.com/WebAPI"

getWebApiVersion(baseUrl)
```


Take a look at the cohorts we created in Atlas.

```{r}
cohorts <- getCohortDefinitionsMetaData(baseUrl)
cohorts
```


Download a cohort definition from Atlas.


```{r}
cohort_def <- getCohortDefinition(cohortId = 21, baseUrl)
str(cohort_def, max.level = 1)
```

Save and reload a cohort definition.

```{r}
jsonlite::write_json(cohort_def, "cohort.json", pretty = T)
# take a look at the cohort.json file

cohort_def <- jsonlite::read_json("cohort.json", simplifyVector = T)
str(cohort_def, max.level = 1)
```

We can also download cohort definitions from OHDSI study packages as well.

```{r}
# download heart disease cohort definition from legend study
download.file("https://raw.githubusercontent.com/OHDSI/Legend/master/inst/cohorts/Coronary%20heart%20disease.json", "Coronary heart disease.json")


chd_def <- jsonlite::read_json("Coronary heart disease.json")
str(chd_def, max.level = 1)

```

Note that this looks a little different. This cohort definition contains just the expression and no information about who created it or when it was created.

```{r}
str(cohort_def$expression, max.level = 1)
```

# Get the SQL for that cohort

Now let's take the json and get the SQL code that will *generate* the list of patient ids, cohort entry, and cohort exit dates.

```{r}
chd_sql <- getCohortSql(chd_def, baseUrl)
cat(substr(chd_sql, 1, 200))
```


# Upload a cohort definition to Atlas
We can also programatically upload cohort cohort definitions to Atlas which you might want to do after setting up Atlas for the first time. Instead of creating all your cohorts from scratch you can simply batch import large numbers of pre-defined and tested cohorts from OHDSI studies like LEGEND.

```{r, eval=F}
postCohortDefinition(name = "[test] Coronary Heart Disease",
                     cohortDefinition = chd_def,
                     baseUrl = baseUrl)
```

Uploading multiple cohorts is very similar to uploading one cohort. In this example we will upload all of the LEGEND cohorts to Atlas in a few lines of code.
```{r, eval=F}

# purrr::walk is similar to apply but is call only for it's action (side effect) and does not return a value
# here we just upload a 3 cohorts but you can try uploading them all by removing [1:3]
purrr::walk(list.files("LEGEND_cohorts")[1:3], function(filename) {
  name <- paste("[LEGEND]", stringr::str_remove(filename, ".json"))
  cohortDefinition <- jsonlite::read_json(paste0("LEGEND_cohorts/", filename))
  postCohortDefinition(name, cohortDefinition, baseUrl)
})

# Note that cohort names must be unique so you might want to include your username in the cohort name to avoid name clashes.

# If you see Cohort: already exists error then you will need to choose a different name prefix (change [LEGEND] to [Your initials or username])

```



# Generate cohorts programmatically from R

After we have uploaded those cohorts we may want to **generate** them on one or more datasets programatically. If we have a large number of cohorts this could save a lot of time.

First lets get the source keys for each CDM database.

```{r}
ROhdsiWebApi::getCdmSources(baseUrl)
```

Generating a single cohort on a single CDM is just one command. First let's find a cohort to generate.

```{r}
library(dplyr)

cohorts <- getCohortDefinitionsMetaData(baseUrl)

cohorts %>% 
  filter(stringr::str_detect(name, "LEGEND")) %>% 
  select(id, name)
```


```{r, eval=F}
ROhdsiWebApi::invokeCohortGeneration(cohorts[1,]$id, baseUrl, sourceKey = "synthea1k")
```


```{r, eval=F}
# generate multiple cohorts on a CDM 
cohortMetaData <- getCohortDefinitionsMetaData(baseUrl) %>% 
  filter(stringr::str_detect(name, "LEGEND")) %>% 
  select(id, name)

purrr::walk(cohortMetaData$id, ~invokeCohortGeneration(cohortId = ., baseUrl = baseUrl, sourceKey = "synthea1k"))
```


```{r, eval=F}

# generate multiple cohorts on a multiple CDMs 

source_key <- c("synthea1k", "synthea100k", "synthea23m")

cohortMetaData <- getCohortDefinitionsMetaData(baseUrl) %>% 
  filter(stringr::str_detect(name, "LEGEND")) %>% 
  select(id, name) %>% 
  slice(1:3) %>% 
  tidyr::expand_grid(source_key)

# call the invokeCohortGeneration function for each combination of cohortId and sourceKey
purrr::walk2(cohortMetaData$id, cohortMetaData$source_key, 
             ~invokeCohortGeneration(cohortId = .x, baseUrl = baseUrl, sourceKey = .y))

```


With CirceR we can also generate human readable cohort descriptions from json files.

```{r}
cohortJson <- paste(readLines("LEGEND_cohorts/Abdominal pain.json"),collapse="\n")
cohortMarkdown <- CirceR::cohortPrintFriendly(cohortJson)

cat(cohortMarkdown)
```


# Cohort Diagnostics

The [CohortDiagnostics](https://github.com/OHDSI/CohortDiagnostics) package provides powerful tools for assessing the quality of a cohort definion and "fitness for use" of a particular dataset for a particular study. 



```{r, eval=F}
library(CohortDiagnostics)

connectionDetails <- createConnectionDetails(dbms = "redshift",
                                             user = "master",
                                             server = Sys.getenv("DB_SERVER"),
                                             password = Sys.getenv("DB_PASSWORD"),
                                             # password = keyring::key_get(),
                                             # password = rstudioapi::askForPassword(),
                                             port = "5439")



cohortMetadata <- getCohortDefinitionsMetaData(baseUrl)

cohortSetReference <- cohortMetadata %>% 
  filter(stringr::str_detect(name, "Legend")) %>% 
  mutate(atlasId = id,
         atlasName = name,
         cohortId = id,
         name = stringr::str_remove_all(name, "Legend|\\W")) %>% 
  select(atlasId, atlasName, cohortId, name)

CohortDiagnostics::runCohortDiagnostics(baseUrl = baseUrl,
                                        connectionDetails = connectionDetails,
                                        cdmDatabaseSchema = "mycdm.synthea100k",
                                        cohortDatabaseSchema = "mycdm.synthea100kresults",
                                        cohortTable = "cohort", 
                                        cohortSetReference = cohortSetReference,
                                        exportFolder = "~/cohortDiagnosticsExport",
                                        runInclusionStatistics = F,
                                        databaseId = "synthea100k")
```


Cohort Diagnosics can also run in "package" mode. This is helpful when you want to run an OHDSI study. We will be discussing OHDSI studies in the next section.

# Exercise
Create a new cohort in Atlas. Then run Cohort Diagnostics on that cohort using R. 

